---
title: "Homework, Week 6"
author: "Joy Payton"
date: "October 3, 2015"
output: 
  html_document: 
    toc: true
    theme: cerulean
---

##Assignment:
- Read the information from an untidy .CSV file into R
- Use tidyr and dplyr as needed to tidy and transform the data.
- Perform analysis to compare the arrival delays for the two airlines.

##Planning:

###Starting format:

We begin with a table that looks like this (some cities left out for brevity, dots represent blank leading cells):

.         |           | Los Angeles | Phoenix | San Diego 
--------- | --------- | ----------- | ------- | -------------
ALASKA    | on time   | 497         | 221     | 212 
.         | delayed   | 62          | 12      | 20
.         |           |             |         | 
AM WEST   | on time   | 694         | 4,840   | 383
.         | delayed   | 117         | 415     | 65

###Final format: 
We begin with the end in mind.  We will want a table that looks like this, instead:

Airline | City | On Time | Delayed
------- | ---- | ------- | -------
Alaska  | Los Angeles | 497 | 62
Alaska  | Phoenix | 221 | 12
Alaska  | San Diego | 212 | 20
Am West | Los Angeles | 694 | 117
Am West | Phoenix | 4840 | 415
Am West | San Diego | 383 | 65

This kind of table can be analyzed easily for on-time performance by city or by airline.

###Assumptions:
We assume that this is an example table and that the actual table could be wider (more cities) or longer (more airlines).  The code in this R Markdown is intended to work for any length of data that follows the same format as the example given and read in.
 
##Load up required packages:
```{r, eval=TRUE}
library(dplyr)
library(tidyr)
```

##Download CSV and read it into an R object

```{r, eval=TRUE}
download.file("https://raw.githubusercontent.com/pm0kjp/IS607/df95386b98f2605b63947b41d670d20c1a9b9e94/flight_ontime_data.csv", "flight_ontime_data.csv", method="curl")
flight_data<-read.csv("flight_ontime_data.csv")
head(flight_data)
```

## Basic, Initial Cleanup

Keep in mind that this is an example.  We want to avoid referencing specific rows / columns or doing manipulation that would be difficult or time consuming if the table were instead to include 40 airlines and 75 cities.

Remove any blank lines (instead of just removing the third row, I tried to do this in such a way that it would work for any length of table that continues the data in the same format):

```{r, eval=TRUE}
flight_data<-filter(flight_data, X.1!="")
head(flight_data)
```

Now I have an "every other row" named / nameless problem in my first column.  This prevents me from using spread() effectively. I'll solve this by first coming up with a subset of rows with names and a subset of rows without names.  I do this in such a way that regardless of the number of rows, this will work:
```{r, eval=TRUE}
name_rows<-seq(from = 1, to = nrow(flight_data), by=2)
nameless_rows<-seq(from = 2, to = nrow(flight_data), by=2)
```

Now I simply copy the corresponding names from the named to the unnamed rows.
```{r, eval=TRUE}
flight_data[nameless_rows,1]<-flight_data[name_rows,1]
head(flight_data)
```

## Using gather to make city into a variable

As shown in my desired final format, I want city to be a variable, instead of being a value spread in the header.  I'll use gather to do this.  As always, I don't hard code in the length but refer to the actual dimension of the data frame so that this code is portable to larger datasets with the same format.

```{r, eval=TRUE}
flight_data<-gather(flight_data, "city", "n", 3:ncol(flight_data))
head(flight_data)
```

## Using spread to make on time / delayed into two separate columns

Again, I referenced my desired output table, and look at the second column (still named X.1).  I want to make these into two columns that will each contain the corresponding count.

```{r, eval=TRUE}
flight_data<-spread(flight_data, X.1, n)
head(flight_data)
```

## Final cleanup

We've got a couple of small updates to do:  We need to name the first column, remove dots from the city names, and replace the space in "on time" with an underscore:
```{r, eval=TRUE}
colnames(flight_data)[1]<-"airline"
flight_data$city<-gsub("\\."," ",flight_data$city)
colnames(flight_data)[4]<-"on_time"
flight_data
```

## Analysis

We are asked to do on-time performance analysis. First, let's summarize overall on time performance (otp) across all cities and airlines:
```{r, eval=TRUE}
summarise(flight_data, on_time_total=sum(on_time), delayed_total=sum(delayed), otp=on_time_total/(on_time_total+delayed_total)*100)
```

We can also use group_by and summarise together to summarize overall performance by airline:
```{r, eval=TRUE}
flight_data %>% group_by(airline) %>% summarise(on_time_total=sum(on_time), delayed_total=sum(delayed), otp=on_time_total/(on_time_total+delayed_total)*100)
```

Or, by city:
```{r, eval=TRUE}
flight_data %>% group_by(city) %>% summarise(on_time_total=sum(on_time), delayed_total=sum(delayed), otp=on_time_total/(on_time_total+delayed_total)*100)
```

Or, by both city and airline:
```{r, eval=TRUE}
flight_data %>% group_by(city, airline) %>% summarise(on_time_total=sum(on_time), delayed_total=sum(delayed), otp=on_time_total/(on_time_total+delayed_total)*100)
```

## Visualization

What about some data visualization?  We could start really simple, with qplot.  We can plot the on time performance (on time divided by all flights) on a per-city basis, with each airline having a color.

```{r, eval=TRUE}
library(ggplot2)
qplot(city,on_time/(on_time+delayed), data=flight_data, color=airline)
```

Well, this helps us see visually that Alaska has a better on-time performance in each market, compared to Am. West, but the plot itself is unattractive.  The y-axis label is ugly, the dots are small, and the plot isn't titled.  Let's do a grouped bar plot, and update the y axis and legend:

```{r, eval=TRUE}
ggplot(flight_data, aes(x=factor(city), fill=factor(airline), y=on_time/(on_time+delayed)*100)) + geom_bar(position="dodge", stat="identity") + ylab("Percent On-Time Performance") + scale_y_continuous(breaks = seq(0, 100, by = 5)) + xlab ("") + scale_fill_discrete(name="Airline")
```
