---
title: "Week 4 Homework, IS 607"
author: "Joy Payton"
date: "September 19, 2015"
output: html_document
---

This file represents partial completion of the IS 607 course in the MS in Data Analytics
Program at the City University of New York.  Note this R Markdown file includes characters
that will be rendered properly only using UTF-8.  If given an encoding option, please
use UTF-8.

```{r, eval=TRUE}
library(stringr)
```

```
We are asked to 
4. Describe the types of strings that conform to the following regular expressions and
construct an example that is matched by the regular expression.
(a) [0-9]+\\$
(b) \\b[a-z]{1,4}\\b
(c) .*?\\.txt$
(d) \\d{2}/\\d{2}/\\d{4}
(e) <(.+?)>.+?</\\1>
```
```
(a) [0-9]+\\$
```
In (a), we see that the bracketed group contains all digits, and is quantified using the plus
sign, so we know that a matching group will have one or more digits.  The double backslash that 
follows indicates a literal single backslash in the pattern text, such that the R interpreter 
will understand that the pattern desired is \\$, or a literal dollar sign. Note that this 
explanation for the double backslash applies broadly in R and will not be revisited in this 
assignment.
**The regex pattern in (a) will match numerical digits (1 or more) followed by a dollar sign.**

We test with a series of strings we believe *will* match the regex for (a).
The first three we expect to match entirely, while the next three will have a matching 
text within the string.

```{r eval=TRUE}
matches<-c("1127$",    # we believe will match entirely
           "0$",       # we believe will match entirely
           "9091118$", # we believe will match entirely
           "price: 1234$ or equiv in Euros",   # we believe will match chunk of text
           "19.176$",                          # we believe will match chunk of text
           "00005$_1712$%")                    # we believe will match chunk of text
str_extract(matches, "[0-9]+\\$")
```

We'll also check for things we expect to NOT match:
```{r eval=TRUE}
non_matches<-c("1127 $",    # we believe will not match (space)
           "$125",          # we believe will not match (inversion of digits and $)
           "9091118",       # we believe will not match (lack of $)
           "1234 dollars",  # we believe will not match (lack of $)
           "ten$",          # we believe will not match (lack of digits)
           "1236\\$")       # we believe will not match: misunderstanding of  \\$
str_extract(non_matches, "[0-9]+\\$")
```

```
(b) \\b[a-z]{1,4}\\b
```
In (b) we begin with a word boundary followed by any 1, 2, 3, or 4 lowercase letter(s).  This 
is followed by another word boundary.  Since we treat of alphabetic characters enclosed by 
word boundaries, we could also describe this pattern as **a 1-4 character long "word" (whether
lexical or not) composed only of lowercase alphabetic characters.** Accented characters would not 
count.

We test with a series of strings we believe *will* match the regex for (b).
The first three we expect to match entirely, while the next three will have a matching 
text within the string.

```{r eval=TRUE}
matches<-c("flop",    # we believe will match entirely
           "zz",      # we believe will match entirely
           "a",       # we believe will match entirely
           "  able 1",   # we believe will match chunk of text
           "wing, ",     # we believe will match chunk of text
           "123: nope!")                    # we believe will match chunk of text
str_extract(matches, "\\b[a-z]{1,4}\\b")
```

We'll also check for things we expect to NOT match:
```{r eval=TRUE}
non_matches<-c("Why?",      # we believe will not match (capital letter)
           "há",            # we believe will not match (accented text)
           "9091118",       # we believe will not match (no alphas)
           "bananas?",      # we believe will not match (too long)
           "fiver",         # we believe will not match (too long)
           "fl1r")          # we believe will not match (digit in word)
str_extract(non_matches, "\\b[a-z]{1,4}\\b")
```

```
(c) .*?\\.txt$
```
In (c) we see **a non-greedy qualifier (?) appended to the "zero or more of any character" 
regex and followed by a .txt and immediately afterwards by the end of the line.**  
This means that R will select the minimum length of text that still matches for the first 
part of the regex, rather important since .* can be greedy indeed and match text that 
was intended to match other more specific regex later in the pattern.

We test with a series of strings we believe *will* match the regex for (c).
The first three we expect to match entirely, while the next three will have a matching 
text within the string.

```{r eval=TRUE}
matches<-c("h12368_dzdjh.co.january.txt",    # we believe will match entirely
           "n1 fakdl    12  .txt",           # we believe will match entirely
           ".txt",                           # we believe will match entirely
           "ring_of_\n  fire.txt",           # we believe will match chunk of text
           "i.jpg don't.jpg  
           know.txt",      # we believe will match chunk of text
           "123: \nnope!.txt")               # we believe will match chunk of text
str_extract(matches, ".*?\\.txt$")
```

We'll also check for things we expect to NOT match:
```{r eval=TRUE}
non_matches<-c("Why.txt ",      # we believe will not match (.txt followed by space)
           "txt",               # we believe will not match (no "dot")
           "9091118.
           txt",                # we believe will not match 
                                # (unmatching, unprintable character between . and text )
           "text.txt1",         # we believe will not match (.txt followed by number, not EOL)
           "five_txt",          # we believe will not match (underscore instead of dot)
           "fl1r")              # we believe will not match (just no)
str_extract(non_matches, ".*?\\.txt$")
```

```
(d) \\d{2}/\\d{2}/\\d{4}
```
The regex for d is quite simple: **two digits followed by a forward slash, followed by an
additional two digits and a slash, and finishing with four digits.**  This is a typical
manner of displaying dates in many written languages.

We test with a series of strings we believe *will* match the regex for (d).
The first three we expect to match entirely, while the next three will have a matching 
text within the string.

```{r eval=TRUE}
matches<-c("00/00/0000",    # we believe will match entirely
           "99/23/1269",    # we believe will match entirely
           "12/25/1977",    # we believe will match entirely
           "dob: 07/01/1980",        # we believe will match chunk of text
           "04/12/1900: never",      # we believe will match chunk of text
           "12312/12/12579")         # we believe will match chunk of text
str_extract(matches, "\\d{2}/\\d{2}/\\d{4}")
```

We'll also check for things we expect to NOT match:
```{r eval=TRUE}
non_matches<-c("00-00-0000",     # we believe will not match (dash instead of slash)
           "3/15/2001",          # we believe will not match (short a digit)
           "12980/1/01970",      # we believe will not match (short a digit)
           "Jan/04/1616",        # we believe will not match (alpha instead of numbers)
           "05 / 12 / 2000",     # we believe will not match (extra spaces)
           "1995/04/12")         # we believe will not match (wrong number of numerals)
str_extract(non_matches, "\\d{2}/\\d{2}/\\d{4}")
```
```
(e) <(.+?)>.+?</\\1>
```
Section (e) features a pattern that will match **an open triangle bracket, some characters (at 
least one), and a closing triangle bracket followed by at least one character and a similar 
bracketed text which matches the first (\\1) with the exception of a leading forward slash.** This
matches the markup of html and uses regex backreferencing.

We test with a series of strings we believe *will* match the regex for (e).
The first three we expect to match entirely, while the next three will have a matching 
text within the string.

```{r eval=TRUE}
matches<-c("<title>Bob</title>",    # we believe will match entirely
           "<img> </img>",          # we believe will match entirely
           "<abc>abc</abc>",        # we believe will match entirely
           "the end.<div> </div> ", # we believe will match chunk of text
           "<a>.</a><b>123",        # we believe will match chunk of text
           "not<b> certain</b>")        # we believe will match chunk of text
str_extract(matches, "<(.+?)>.+?</\\1>")
```

We'll also check for things we expect to NOT match:
```{r eval=TRUE}
non_matches<-c("<abc>abc</ac>",      # we believe will not match (closing tag does not match)
           "</br>",                  # we believe will not match (self closing tag with no opening tag)
           "<div id='bob'> </div>",  # we believe will not match (closing tag doesn't match)
           "> the end </img>",       # we believe will not match (broken opening tag)
           "<p></p>",                # we believe will not match (no content between tags)
           "<html>")                 # we believe will not match (only opening tag)
str_extract(non_matches, "<(.+?)>.+?</\\1>")
```
```
5.  Rewrite the expression [0-9]+\\$ in a way that all elements are altered but the
expression performs the same task.
```
Original  |  Replacement
--------- | -------------
[0-9]     | [[:digit]]
+         | {1,}
\\$       | \\x24
[0-9]+\\$ | [[:digit:]]{1,}\\x24
We can check this replacement table (the last replacement is a hex representation of the $ character)
```{r eval=TRUE}
matches<-c("1127$",    # we believe will match entirely
           "0$",       # we believe will match entirely
           "9091118$", # we believe will match entirely
           "price: 1234$ or equiv in Euros",   # we believe will match chunk of text
           "19.176$",                          # we believe will match chunk of text
           "00005$_1712$%")                    # we believe will match chunk of text
str_extract(matches, "[[:digit:]]{1,}\\x24")
```

We'll also check for things we expect to NOT match:
```{r eval=TRUE}
non_matches<-c("1127 $",    # we believe will not match (space)
           "$125",          # we believe will not match (inversion of digits and $)
           "9091118",       # we believe will not match (lack of $)
           "1234 dollars",  # we believe will not match (lack of $)
           "ten$",          # we believe will not match (lack of digits)
           "1236\\$")       # we believe will not match: misunderstanding of  \\$
str_extract(non_matches, "[[:digit:]]{1,}\\x24")
```

```
6. Consider the mail address chunkylover53[at]aol[dot]com.
(a) Transform the string to a standard mail format using regular expressions.
(b) Imagine we are trying to extract the digits in the mail address.  To do so we write
the expression [:digit:].  Explain why this fails and correct the expression.
(c) Instead of using the predefined character classes, we would like to use the predefined
symbols to extract the digits in the mail address.  To do so we write the expression
\\D.  Explain why this fails and correct the expression.
```
In part (a) we want to change [at] to @ and [dot] to . .  We'll start with our replacement 
patterns:
```{r eval=TRUE}
at_pattern<-"\\[at\\]"
dot_pattern<-"\\[dot\\]"
```

Now we replace matching text with what's preferred:
```{r eval=TRUE}
better_email_address<-str_replace("chunkylover53[at]aol[dot]com", at_pattern, '@')
better_email_address<-str_replace(better_email_address, dot_pattern, '.')
better_email_address
```

In part (b) we forget that the character class needs a quantifier to allow for digit 
strings of length >1.  We should add a plus, unless we're using a function that 
extracts all.  Additionally, we should add an additional layer of square brackets to 
allow for proper interpretation by the R regex interpreter (although in my
Mac OS environment and R version, the expected confusion doesn't occur).

Consider the difference:
```{r eval=TRUE}
str_extract(better_email_address, "[:digit:]")
str_extract(better_email_address, "[[:digit:]]+")
```

In part (c) we forget that capitalization matters.  We should use a lower case \\\\d
instead of an upper case, which captures the converse!
Consider the difference:
```{r eval=TRUE}
str_extract("chunkylover53[at]aol[dot]com", "\\D+")
str_extract("chunkylover53[at]aol[dot]com", "\\d+")
```
